### 2024-11-22 teaching/math240f24/draft.tex
lecture 21 nov

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
graphs}{74}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 15.\kern .5em Eulerian
trails and circuits}{81}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx IV. Combinatorics \kern 3pt}}{88}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 17.\kern .5em Basic counting techniques}{89}}

@@

Now suppose that there is some component $G_i$ that is not bipartite.
Now let $V = A\cup B$
be any partition of $V$ into disjoint nonempty sets. Then $A_i = A\cap V_i$
and $B_i = B\cap V_i$ form a partition of $V_i$
into two nonempty sets. But $G_i$ is not bipartite, so there is some $e\in E$
that either has both endpoints in $A_i$ or both endpoints in $B_i$. This means
that $G$ is not bipartite either.\slug

@@

Suppose we have a pool of $100$ hockey players and we want to dress $20$
of them for a game, giving them the numbers $1$ through $20$. How many ways are there to do this?
Well, there are $100$ choices for the player with jersey number $1$, then after
picking that player, there are $99$ choices for the player with jersey number $2$, and so on,
so the total number is
$$100\cdot 99\cdot 98 \cdots 82\cdot 81 = {100!\over 80!}.$$
\bye

\noindent\datestamp{23}{} \hskip\parindent
In general, we define a {\it $k$-permutation} of a set $X$ to be an ordered subset
of $X$ (i.e., the order of the selection matters) of size $k$. If $|X| = n$,
the number of $k$-permutations of $X$ is denoted $P(n,k)$ and equals
$$ P(n,k) = n\cdot (n-1)\cdots (n-k+1) = {n!\over (n-k)!}.$$
This number is also sometimes called the {\it falling factorial} and denoted $n^{\underline k}$.
It is the same as the number of functions from $\{1,2\ldots, k\}$ to $X$ that are injective.
(Pause a moment to think about it if you need to.) Recall that if $|S| = s$ and $|T| = t$,
the total cardinality of $S^T$ is $s^t$. Well, the number of
{\it injective} functions from $T$ to $S$ is $P(s,t) = s^{\underline{t}}$.

@@

\medskip\boldlabel\datestamp{24}{} Identities involving binomial coefficients.
We already observed earlier that the number of ways of picking $k$ objects out of a
set of size $n$ is the same as the number of picking $n-k$ objects out of the same set.
This can be seen algebraically by expanding
$${n\choose k} = {n!\over k!(n-k)!} = {n!\over (n-k)!\bigl(n-(n-k)\bigr)!} = {n\choose n-k}.$$
Here is another identity involving binomial coefficients.

@@

Another thing we notice about Pascal's triangle is that the first row sums to $1$, the second
row sums to $2$, the third to $4$, the fourth to $8$, and so on. We might therefore
conjecture that the $n$th row of Pascal's triangle should always sum up to $2^n$. Indeed, we
have the following easy corollary of the binomial theorem.

@@

Let $1\le k\le p-1$ and expand
$${p\choose k} = {p!\over k!(p-k)!}$$
Multiplying both sides by $k!(p-k)!$, we get
$$k!(p-k)!{p\choose k} = p! = p(p-1)!.$$
This is an integer and the
right-hand side is a multiple of $p$, so the left-hand side is as well. Then since $p$ divides
the product $k!(p-k)!{p\choose k}$, either $p$ divides $k!(p-k)!$ or it divides $p\choose k$.
But
$$k!(p-k)! = k(k-1)\cdots 2\cdot 1\cdot (p-k)(p-k-1)\cdots 2\cdot 1,$$
and all of the factors on the right-hand side are positive integers less than $p$. So $p$ cannot divide
any of them, and hence $p$ does not divide $k!(p-k)!$. We conclude that $p\divides {p\choose k}$.\slug

\medskip\boldlabel A day spent selling milk.
The binomial theorem can be used to find the value of a specific coefficient in a polynomial,
without necessarily expanding the whole thing out. To illustrate this, consider the following
scenario. Imagine that a farmer runs a roadside
milk stand. Each time a customer visits they either
\medskip
\item{i)} pay her $\$6$ and buy one bottle of white, chocolate, or strawberry milk; or
\smallskip
\item{ii)} sell her back a glass bottle for $\$1$.
\medskip
So after one customer visits, there are three ways in which the farmer could have
made $\$6$, and one way in which she could have made $-\$1$. Let us arrange these
possibilities as a function of $x$; we write $3x^6 + x^{-1}$.

>>>>

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
graphs}{74}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 15.\kern .5em Eulerian
trails and circuits}{81}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx IV. Combinatorics \kern 3pt}}{88}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 17.\kern .5em Basic counting techniques}{89}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 18.\kern .5em Permutations and
combinations}{92}}

@@

Now suppose that there is some component $G_i$ that is not bipartite.
Now let $V = A\cup B$
be any partition of $V$ into disjoint nonempty sets. Then $A_i = A\cap V_i$
and $B_i = B\cap V_i$ form a partition of $V_i$
into two nonempty sets. But $G_i$ is not bipartite, so there is some $e\in E$
that either has both endpoints in $A_i$ or both endpoints in $B_i$. This means
that $A\cup B$ is not a valid bipartition, and since $A$ and $B$ were arbitrary,
we conclude that
$G$ is not bipartite either.\slug

@@

Suppose we have a pool of $100$ hockey players and we want to dress $20$
of them for a game, giving them the numbers $1$ through $20$. How many ways are there to do this?
Well, there are $100$ choices for the player with jersey number $1$, then after
picking that player, there are $99$ choices for the player with jersey number $2$, and so on,
so the total number is
$$100\cdot 99\cdot 98 \cdots 82\cdot 81 = {100!\over 80!}.$$

\noindent\datestamp{21}{xi} \hskip\parindent
In general, we define a {\it $k$-permutation} of a set $X$ to be an ordered subset
of $X$ (i.e., the order of the selection matters) of size $k$. If $|X| = n$,
the number of $k$-permutations of $X$ is denoted $P(n,k)$ and equals
$$ P(n,k) = n\cdot (n-1)\cdots (n-k+1) = {n!\over (n-k)!}.$$
This number is also sometimes called the {\it falling factorial} and denoted $n^{\underline k}$.
It is the same as the number of functions from $\{1,2\ldots, k\}$ to $X$ that are injective.
(Pause a moment to think about it if you need to.) Recall that if $|S| = s$ and $|T| = t$,
the total cardinality of $S^T$ is $s^t$. Well, the number of
{\it injective} functions from $T$ to $S$ is $P(s,t) = s^{\underline{t}}$.

@@

\medskip\boldlabel Identities involving binomial coefficients.
We already observed earlier that the number of ways of picking $k$ objects out of a
set of size $n$ is the same as the number of picking $n-k$ objects out of the same set.
This can be seen algebraically by expanding
$${n\choose k} = {n!\over k!(n-k)!} = {n!\over (n-k)!\bigl(n-(n-k)\bigr)!} = {n\choose n-k}.$$
Here is another identity involving binomial coefficients.

@@

The only number that appears infinitely many times is $1$, since any $n\in \ZZ$
can only appear in the first $n+1$ rows of the triangle (if $m>n$, then for all
$0\le k\le m$, ${m\choose k}$ is either $1$ or at least $m$). A fun conjecture to think
about is Singmaster's conjecture, put forth by D.~Singmaster in 1971: {\sl There
exists some number $C$ such that every integer $n\ge 2$ appears in Pascal's triangle
at most $C$ times.} Whether this statement is true is still an open problem.
The integer (besides $1$) known to appear most often in Pascal's triangle is
$3003$. It appears eight times, and no other integer is known to appear eight times.

But let us not be too diverted by lofty conjectures. Returning to things we can prove,
another thing we notice about Pascal's triangle is that the first row sums to $1$, the second
row sums to $2$, the third to $4$, the fourth to $8$, and so on. We might therefore
conjecture that the $n$th row of Pascal's triangle should always sum up to $2^n$. Indeed, we
have the following easy corollary of the binomial theorem.

@@

Let $1\le k\le p-1$ and expand
$${p\choose k} = {p!\over k!(p-k)!}$$
Multiplying both sides by $k!(p-k)!$, we get
$$k!(p-k)!{p\choose k} = p! = p(p-1)!.$$
This is an integer and the
right-hand side is a multiple of $p$, so the left-hand side is as well. Then since $p$ divides
the product $k!(p-k)!{p\choose k}$, either $p$ divides $k!(p-k)!$ or it divides $p\choose k$.
But
$$k!(p-k)! = k(k-1)\cdots 2\cdot 1\cdot (p-k)(p-k-1)\cdots 2\cdot 1,$$
and all of the factors on the right-hand side are positive integers less than $p$. So $p$ cannot divide
any of them, and hence $p$ does not divide $k!(p-k)!$. We conclude that $p\divides {p\choose k}$.\slug
\bye

\medskip\boldlabel\datestamp{24}{} A day spent selling milk.
The binomial theorem can be used to find the value of a specific coefficient in a polynomial,
without necessarily expanding the whole thing out. To illustrate this, consider the following
scenario. Imagine that a farmer runs a roadside
milk stand. Each time a customer visits they either
\medskip
\item{i)} pay her $\$6$ and buy one bottle of white, chocolate, or strawberry milk; or
\smallskip
\item{ii)} sell her back a glass bottle for $\$1$.
\medskip
So after one customer visits, there are three ways in which the farmer could have
made $\$6$, and one way in which she could have made $-\$1$. Let us arrange these
possibilities as a function of $x$; we write $3x^6 + x^{-1}$.
